Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-07 doc/invt.tex
improving figure errscansig.gp

Initially,
we set the bias potential to zero,
$v_i \equiv 0$,
%
and equilibrate the system under
a constant updating magnitude,
$\alpha_0$.
%
After this,
we reset the origin of the time, $t$,
and started the target schedule.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsbin.pdf}
  }
  \caption{
    \label{fig:errsbin}
    Error, $E$, versus the
    relative updating magnitude,
    %proportionality constant,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the single-bin (WL) scheme.
    %
    The results have been averaged over $1000$ independent runs.
    %
    \note{The figure is produced by \texttt{doc/fig/errsbin.gp}.
      The data are produced by the script \texttt{invt2run.py},
      and saved under \texttt{data/singlebin}
      (see the \texttt{Makefile} there).
    }
  }
\end{center}
\end{figure}

@@

As an example of the multiple-bin schemes,
we chose the nearest-neighbor updating scheme
described in Sec. \eqref{sec:nnscheme}.
%
We again find good agreement
between simulation and analytical results.
%
A new feature is that the optimal $\lambda$
depends on the simulation length, $T$.
%
As shown in Fig. \ref{fig:err_nnbr},
the normalized error, $(T + t_0) \, \Err$,
hence the optimal value of $1/\lambda$,
depended non-trivially on the simulation length, $T$,
in the global sampling case.
%
The optimal $1/\lambda$ gradually increased
with simulation length,
which parallel the gradual shift of
the instantaneous eigenvalue $\lambda(t)$
as discussed in Sec. \ref{sec:optschedule}.
%which is in agreement with
%the discussion in Sec. \ref{sec:optlambda}.
%
For the local sampling case,
the $T$-dependence was hardly visible
because of the predominantly large $\Gamma_k$'s
of the small $k$ modes.
%
The optimal $\lambda$ remained around $1.0$,
as suggested by Eq. \eqref{eq:lambda_nn_onestep}.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errnnbr.pdf}
  }
  \caption{
    \label{fig:err_nnbr}
    Normalized error, $(T + t_0) \, E$,
    versus the %proportionality constant,
    relative updating magnitude,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the nearest-neighbor updating scheme with $\mu_1 = 0.24$.
    %
    The results have been averaged over $1000$ independent runs.
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/optacmp.pdf}
  }
  \caption{
    \label{fig:optacmp}
    Optimal schedules from Eq. \eqref{eq:q_opt},
    versus the optimized inverse-time schedule
    Eq. \eqref{eq:alpha_invtlambda}
    for the Gaussian updating scheme
    with $\sigma = 10$.
  }
\end{center}
\end{figure}

As shown in Fig. \ref{fig:optacmp}(a),
the optimal schedule has a more complex curvature
than the optimized inverse-time schedule.
%
In Fig. \ref{fig:optacmp}(b),
we show that in the former case,
the rate of change of $\alpha^{-1}(t)$
increases gradually
from $\lambda(0)$ to $\lambda(T) \approx 1$,
as discussed in Sec. \ref{sec:optschedule}.
%
The errors in the two cases were
$2.7 \times 10^{-7}$
and
$5.9 \times 10^{-7}$,
respectively,
for the global sampling process,
%
or
$9.1 \times 10^{-5}$
and
$1.67 \times 10^{-4}$,
respectively
for the local sampling process,
suggesting a slight edge of
the optimal schedule from Eq. \eqref{eq:q_opt}.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/sinc.pdf}
  }
  \caption{
    \label{fig:sinc}
    Bandpass updating schemes.
    (a) Optimal schedule for a periodic variable
    with $K = 3$,
    and that for a non-periodic variable
    with $K = 6$.
    (b) Updating kernels
    compared to a Gaussian with a similar width,
    $\sigma = 5$.
    The lines are a guide to the eyes.
    Inset. Updating matrix in the non-periodic case
    is free of the zigzaggedness shown in the kernel.
  }
\end{center}
\end{figure}

@@

For a finite-length simulation,
the optimal updating scheme
depends on the simulation length, $T$,
and other parameters.
%
To illustrate this,
we compare Gaussian updating schemes
with different width, $\sigma$.
%
For the global sampling process,
we see from Fig. \ref{fig:errsigscan}(a)
that the minimal error was achieved
by the widest window for the shorter simulation
of $T = 10^8$,
while the minimum around $\sigma = 0$
became more favorable as the simulation lengthens.
%
However,
the shift was hard to detect
for the local sampling scheme,
as shown in Fig. \ref{fig:errsigscan}(b).

In contrast,
the normalized error of the bandpass updating scheme
was insensitive to the simulation length,
in agreement with Eq. \eqref{eq:error_sinc}.
%
Further, in all cases the bandpass scheme
appeared to be superior to the Gaussian updating scheme.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:errsigscan}
    Normalized error, $(T + t_0) \, E$,
    versus the width, $\sigma$,
    for the Gaussian
    and the bandpass updating schemes.
    %
    In the latter case,
    $\sigma$ is given by Eq. \eqref{eq:sigma_equiv}.
  }
\end{center}
\end{figure}

@@

For a finite-length simulation,
we need to include the residual error, $\Err_R$.
%
By Eq. \eqref{eq:alpha_invtlambda}, we have
%
\begin{equation}
  q(T)
  =
  \frac{ 1 } { \lambda }
  \log\left(
    \frac{ T + t_0 } { t_0 }
  \right)
  ,
\notag
%\label{eq:qt_invtlambda}
\end{equation}
%
so that Eq. \eqref{eq:error_res1} becomes
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        t_0   }
  \left(
      \frac{   t_0   }
           { T + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that
the system was initially equilibrated
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql},
as well as Eq. \eqref{eq:t0_sinc}.
%
The total error is
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \frac
  {
    \Gamma_k \, \nu_k \,
    \left[
      \nu_k
      -
      (\nu_k - 1)
      \left(
        \frac{ t_0 } { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
    \right]
  }
  {
    (T + t_0) \, (2 \, \nu_k - 1)
  }
  .
\label{eq:error_invt}
\end{align}
%
%where
%$$
%\Err^f_k
%\equiv
%\left\langle
%  y_k^2
%\right\rangle_{ \alpha(t) }
%=
%\frac{  \Gamma_k \, \nu_k   }
%     {    2 \, (T + t_0)    }
%,
%$$
%is the equilibrium, or saturated, error of mode $k$
%at the terminal updating magnitude
%of $\alpha(T) = 1/[\lambda (T + t_0)]$.

@@

\note{

@@

Let us first rewrite Eq. \eqref{eq:error_invt} as
%
$$
\Err
=
\sum_{ k = 2 }^n
  \frac{ \Gamma_k  }
       { (T + t_0) }
  G\left( \nu_k, \frac{ t_0 } { T + t_0} \right),
$$
%
where the $k$th component is proportional to
%
$$
G\left( \nu_k, r \right)
=
\frac { \nu_k }
      { 2 \, \nu_k - 1 }
\,
\left[
  \nu_k
  -
  (\nu_k - 1) \,
  r^{2 \, \nu_k - 1}
\right].
$$
%
As shown in Fig. \ref{fig:err_component},
the function $G(\nu_k, r)$ depends non-trivially on $\nu_k$.
%
For a fixed and sufficiently small $r = t_0 / (T + t_0)$,
$G(\nu_k, r)$ has two local minima along $\nu_k$,
one at $\nu_k = 0$,
and the other shallower one around $\nu_k = 1$.
%
For a short simulation, or a large $r$,
the minimal error may be achieved at $\nu_k \approx 0$,
corresponding to a large value of $\lambda$
($\lambda \gg \lambda_k$).
%
However, this minimum is unstable,
for $G(\nu_k, r)$ increases rapidly
with decreasing $r$, or increasing simulation length,
under a fixed $\nu_k$.
%
In other words,
asymptotically, the minimal error
is always achieved at the other minimum,
$\nu_k = 1$ or $\lambda = \lambda_k$.
%
In this case,
$G(\nu_k, r) \to G(1, 0) = 2$
and the total error is roughly twice $E_\mathrm{sat}$.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errcomp.pdf}
  }
  \caption{
    \label{fig:err_component}
    Component of the error, $G(\nu_k, r)$,
    versus $\nu_k = \lambda_k / \lambda$,
    where $r = t_0 / (T + t_0)$.
    %
  }
\end{center}
\end{figure}

>>>>

Initially,
we set the bias potential to zero,
$v_i \equiv 0$,
%
and equilibrate the system under
a constant updating magnitude,
$\alpha_0 = 10^{-4}$.
%
After this,
we reset the origin of the time, $t$,
and started the target schedule.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsbin.pdf}
  }
  \caption{
    \label{fig:errsbin}
    Error, $E$, versus the
    relative updating magnitude,
    %proportionality constant,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the single-bin (WL) scheme.
    %
    The results have been averaged over $1000$ independent runs.
    %
    \note{The figure was produced by \texttt{doc/fig/errsbin.gp}.
      The data were produced by the script \texttt{data/invt2run.py},
      and saved under \texttt{data/singlebin}
      (see the \texttt{Makefile} there).
    }
  }
\end{center}
\end{figure}

@@

As an example of the multiple-bin schemes,
we chose the nearest-neighbor updating scheme
described in Sec. \eqref{sec:nnscheme}.
%
We again find good agreement
between simulation and analytical results.
%
A new feature is that the optimal $\lambda$
depends on the simulation length, $T$.
%
As shown in Fig. \ref{fig:errnnbr},
the normalized error, $(T + t_0) \, \Err$,
hence the optimal value of $1/\lambda$,
depended non-trivially on the simulation length, $T$,
in the global sampling case.
%
The optimal $1/\lambda$ gradually increased
with simulation length,
which parallel the gradual shift of
the instantaneous eigenvalue $\lambda(t)$
as discussed in Sec. \ref{sec:optschedule}.
%which is in agreement with
%the discussion in Sec. \ref{sec:optlambda}.
%
For the local sampling case,
the $T$-dependence was hardly visible
because of the predominantly large $\Gamma_k$'s
of the small $k$ modes.
%
The optimal $\lambda$ remained around $1.0$,
as suggested by Eq. \eqref{eq:lambda_nn_onestep}.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errnnbr.pdf}
  }
  \caption{
    \label{fig:errnnbr}
    Normalized error, $(T + t_0) \, E$,
    versus the %proportionality constant,
    relative updating magnitude,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the nearest-neighbor updating scheme with $\mu_1 = 0.24$.
    %
    The results have been averaged over $1000$ independent runs.
    \note{This figure was produced by \texttt{doc/fig/errnnbr.gp}.
      The data were produced by \texttt{data/invt2run.py},
      saved under \texttt{data/nb0.24}
      (see the \texttt{Makefile} there).
    }
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/optacmp.pdf}
  }
  \caption{
    \label{fig:optacmp}
    Optimal schedules from Eq. \eqref{eq:q_opt},
    versus the optimized inverse-time schedule
    Eq. \eqref{eq:alpha_invtlambda}
    for the Gaussian updating scheme
    with $\sigma = 10$.
    \note{This figure was produced by \texttt{doc/fig/optacmp.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h}.
      The data were saved under the directory \texttt{data/opta}.
      See the \texttt{make} object \texttt{sigsingle}
      in the \texttt{Makefile} there.
    }
  }
\end{center}
\end{figure}

As shown in Fig. \ref{fig:optacmp}(a),
the optimal schedule is a more complex curve
than the optimized inverse-time schedule.
%
In Fig. \ref{fig:optacmp}(b),
we show that in the former case,
the rate of change of $\alpha^{-1}(t)$
increases gradually
from $\lambda(0)$ to $\lambda(T) \approx 1$,
as discussed in Sec. \ref{sec:optschedule}.
%
The errors in the two cases were
$2.7 \times 10^{-7}$
and
$6 \times 10^{-7}$,
respectively,
for the global sampling process,
%
or
$9.1 \times 10^{-5}$
and
$1.7 \times 10^{-4}$,
respectively
for the local sampling process,
showing the superiority of
the optimal schedule from Eq. \eqref{eq:q_opt}.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/sinc.pdf}
  }
  \caption{
    \label{fig:sinc}
    Bandpass updating schemes.
    (a) Optimal schedule for a periodic variable
    with $K = 3$,
    and that for a non-periodic variable
    with $K = 6$.
    (b) Updating kernels
    compared to a Gaussian with a similar width,
    $\sigma = 5$.
    The lines are a guide to the eyes.
    Inset. Updating matrix in the non-periodic case
    is free of the zigzaggedness shown in the kernel.
    \note{This figure was produced by \texttt{doc/fig/sinc.gp}.
      The data were computed by the program \texttt{prog/predict},
      see particularly the function \texttt{intq\_save()}
      in \texttt{prog/intq.h} for panel (a),
      and the functions \texttt{savewin()}
      and \texttt{savewinmat()}
      in \texttt{prog/invt.h} for panel (b),
      The data were saved under the directory \texttt{data/sinc}.
      See the \texttt{make} object \texttt{sincsingle}
      in the \texttt{Makefile} there.
    }
  }
\end{center}
\end{figure}

@@

For a finite-length simulation,
the optimal updating scheme
depends on the simulation length, $T$,
and other parameters.
%
To illustrate this,
we compare Gaussian updating schemes
with different width, $\sigma$.
%
For the global sampling process,
we see from Fig. \ref{fig:errsigscan}(a)
that the minimal error was achieved
by the widest window for the shorter simulation
of $T = 10^8$.
%
However, as the simulation lengthens,
the minimum at $\sigma = 0$ became more favorable.
%
However,
the shift was hard to detect
for the local sampling scheme,
as shown in Fig. \ref{fig:errsigscan}(b).

In contrast,
the normalized error of the bandpass updating scheme
was insensitive to the simulation length,
in agreement with Eq. \eqref{eq:error_sinc}.
%
Further, in all cases the bandpass updating scheme
appeared to be superior to the Gaussian one.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:errsigscan}
    Normalized error, $(T + t_0) \, E$,
    versus the width, $\sigma$,
    for the Gaussian
    and the bandpass updating schemes.
    %
    In the latter case,
    $\sigma$ is given by Eq. \eqref{eq:sigma_equiv}.
    %
    Here, we assumed $t_0 \equiv T/(e^{q(T)} - 1)$
    for the optimal value of $q(T)$.
    %
    The points were obtained from averaging results
    from 1000 independent simulations;
    the curves are theoretical predictions.
    %
    \note{This figure was produced by
      \texttt{doc/fig/errsigscan.gp}.
      The data were saved under
      \texttt{data/scan}.
    }
  }
\end{center}
\end{figure}

@@

For a finite-length simulation,
we need to include the residual error, $\Err_R$.
%
By Eq. \eqref{eq:alpha_invtlambda}, we have
%
\begin{equation}
  q(T)
  =
  \frac{ 1 } { \lambda }
  \log\left(
    \frac{ T + t_0 } { t_0 }
  \right)
  ,
\notag
%\label{eq:qt_invtlambda}
\end{equation}
%
so that Eq. \eqref{eq:error_res1} becomes
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        t_0   }
  \left(
      \frac{   t_0   }
           { T + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that
the system was initially equilibrated
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql},
as well as Eq. \eqref{eq:t0_sinc}.
%
The total error is
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \frac
  {
    \Gamma_k \, \nu_k \,
    \left[
      \nu_k
      +
      (\nu_k - 1)
      \left(
        \frac{ t_0 } { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
    \right]
  }
  {
    (T + t_0) \, (2 \, \nu_k - 1)
  }
  .
\label{eq:error_invt}
\end{align}
%
%where
%$$
%\Err^f_k
%\equiv
%\left\langle
%  y_k^2
%\right\rangle_{ \alpha(t) }
%=
%\frac{  \Gamma_k \, \nu_k   }
%     {    2 \, (T + t_0)    }
%,
%$$
%is the equilibrium, or saturated, error of mode $k$
%at the terminal updating magnitude
%of $\alpha(T) = 1/[\lambda (T + t_0)]$.

@@

\note{The following subsection is deleted in the public version.

@@

Let us first rewrite Eq. \eqref{eq:error_invt} as
%
$$
\Err
=
\sum_{ k = 2 }^n
  \frac{ \Gamma_k  }
       { (T + t_0) }
  G\left( \nu_k, \frac{ t_0 } { T + t_0} \right),
$$
%
where the $k$th component is proportional to
%
$$
G\left( \nu_k, r \right)
=
\frac { \nu_k
        \,
        \left[
          \nu_k
          +
          (\nu_k - 1) \,
          r^{2 \, \nu_k - 1}
        \right]
      }
      { 2 \, \nu_k - 1 }
.
$$
%
As shown in Fig. \ref{fig:err_component},
the function $G(\nu_k, r)$ depends non-trivially on $\nu_k$.
%
For a fixed and sufficiently small $r = t_0 / (T + t_0)$,
$G(\nu_k, r)$ has two local minima along $\nu_k$,
one at $\nu_k = 0$,
and the other shallower one around $\nu_k = 1$.
%
For a short simulation, or a large $r$,
the minimal error may be achieved at $\nu_k \approx 0$,
corresponding to a large value of $\lambda$
($\lambda \gg \lambda_k$).
%
However, this minimum is unstable,
for $G(\nu_k, r)$ increases rapidly
with decreasing $r$, or increasing simulation length,
under a fixed $\nu_k$.
%
In other words,
asymptotically, the minimal error
is always achieved at the other minimum,
$\nu_k = 1$ or $\lambda = \lambda_k$.
%
In this case,
$G(\nu_k, r) \to G(1, 0) = 2$
and the total error is roughly twice $E_\mathrm{sat}$.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errcomp.pdf}
  }
  \caption{
    \label{fig:err_component}
    Component of the error, $G(\nu_k, r)$,
    versus $\nu_k = \lambda_k / \lambda$,
    where $r = t_0 / (T + t_0)$.
    %
    \note{This figure is produced by
      \texttt{doc/fig/errcomp.gp}.
    }
  }
\end{center}
\end{figure}
